Calculating the nonequilibrium observable by a speed-boosting technique

### 2014-04-29 writedown/accelerate/main.tex
backup

Now we want to eliminate the fast modes in~\eqref{eqn:sde-2}, so that
the largest time-step is not smaller than the original
dynamics~\eqref{eqn:sde-1}, and at the same time the numerical
accuracy for calculating the observable should be preserved. Please notice
that the numerical accuracy of the trajectories are not necessarily
preserved.
Therefore, we need some estimate on the trajectory, which will
be developed under the framework of  Girsanov  transformation.
Now assume that the original forcing can be split by
\begin{align}
  \vect b(\vect x) = \vect b_s (\vect x) + \vect b_f(\vect x)
\end{align}
where $\vect b_f(\vect x)$ is the fast part of the force, while $\vect b_s
(\vect x)$ is the slow part.
We revise the force of the  accelerated dynamics~\eqref{eqn:sde-2} by
\begin{align}
  \hat{\vect b} = \alpha \vect b_s + \vect b_f,
\end{align}
and keeps the diffusion coefficient $\tilde\sigma$ unchanged.
Therefore, the difference between the observable calculated by the
accelerated dynamics and the original observable is given by 
\begin{align}\nonumber
  O(T) &= \tilde O(T/\alpha)\\\nonumber  
  &=\int \dd\vect y_0 {\rho}(\vect y_0, 0)
  \int_{\mc\{T/\alpha; \vect y_0, 0\}} 
  O[\vect y_t]\,\dd\tilde\pathmeas[\vect y_t]\\\nonumber
  &=
  \int \dd\vect y_0 {\rho}(\vect y_0, 0)
  \int_{\mc\{T/\alpha; \vect y_0, 0\}}
  O[\vect y_t] \frac{\dd\tilde\pathmeas[\vect y_t]}{\dd\hat\pathmeas[\vect y_t]}
  \dd\hat\pathmeas[\vect y_t]
\end{align}
Since we have $\tilde{\vect b} - \hat{\vect b} = (\alpha - 1)\vect b_f$,
by the Girsanov transformation, we have
\begin{align}\label{eqn:meas-reweight}
  \frac{\dd\tilde\pathmeas[\vect y_t]}{\dd\hat\pathmeas[\vect y_t]}
  = \exp
  \Big[
  \frac{\alpha-1}{\tilde{\sigma}}\int_0^{T/\alpha}\vect b_f(\vect y_s)\cdot\dd\vect w_s
  -
  \frac{(\alpha-1)^2}{2\tilde\sigma^2} \int_0^{T/\alpha}\vect b_f^2(\vect y_s)\dd t\,
  \Big]
\end{align}
We therefore have the error estimate
\begin{align}\label{eqn:o-error}
  O(T) = \hat O(T/\alpha)  +
  \int \dd\vect y_0 {\rho}(\vect y_0, 0)\cdot \me(\vect y_0, T/\alpha)
\end{align}
where
\begin{align}
  \me(\vect y_0, t) = 
  \int_{\mc\{T/\alpha; \vect y_0, 0\}}
  O[\vect y_t]
  \cdot
  W[\vect y_t]\,\dd\hat\pathmeas
\end{align}
and $W[\vect y_s]$ is defined by
\begin{align}
  W[\vect y_t] = 
  \exp
  \Big[
  \frac{\alpha-1}{\tilde{\sigma}}\int_0^{T/\alpha}\vect b_f(\vect y_s)\cdot\dd\vect w_s
  -
  \frac{(\alpha-1)^2}{2\tilde\sigma^2} \int_0^{T/\alpha}\vect b_f^2(\vect y_s)\dd t\,
  \Big]
  -1
\end{align}
Now the problem is that when the accelerate
approximation~\eqref{eqn:o-error} has a good accuracy. One possibility
is that the measure reweighting~\eqref{eqn:meas-reweight} is small.
However, as a matter of fact, the reweighting is usually not small,
because the fast part of the force is usually large. The other possibility
is that the observable $O[\vect y]$ has vanished correlation with
$W[\vect y]$.
This is usually true if in the system the fast DOFs are decoupled with
the slow variables, and the observation is only on the space spanned
by the slow variables.

@@

The other fast variable is the repulsive part of the Lennard-Jones
interaction, which can be removed by the following way: we define a
smooth switching function $s(r)$ that has value 0 at repulsive part of
the LBJ potential, while has value of 1 at the attractive part. Then
the accelerated LJ potential (denoted by $\hat u(r)$) can be given by
\begin{align}
  \hat u(r) = \alpha s(r) u(r) + (1 - s(r)) u(r)
\end{align}
where $u(r)$ is the original LJ potential. It is clear that
$\hat u(0) = u(0)$, and $\hat u(\infty) = \alpha u(\infty)$.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-5.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-5.eps}\\
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-2.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-3.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-4.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-5.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-5.eps}\\
  \caption{The integrated probability fluxes. $\alpha = 2,4,8$.}
  \label{fig:tmp3}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1-3.eps}\\
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1-1.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1-2.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1-3.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1-3.eps}
  \caption{The integrated probability fluxes. $\alpha = 4,8$.}
  \label{fig:tmp3}
\end{figure}

>>>>

Now we want to eliminate the fast modes in~\eqref{eqn:sde-2}, so that
the largest time-step is not smaller than the original
dynamics~\eqref{eqn:sde-1}, and at the same time the numerical
accuracy for calculating the observable should be preserved. Please notice
that the numerical accuracy of the trajectories are not necessarily
preserved.
Therefore, we need some estimate on the trajectory, which will
be developed under the framework of  Girsanov  transformation.
Now assume that the original forcing can be split by
\begin{align}
  \vect b(\vect x) = \vect b_s (\vect x) + \vect b_f(\vect x)
\end{align}
where $\vect b_f(\vect x)$ is the fast part of the force, while $\vect b_s
(\vect x)$ is the slow part.
We revise the force of the  accelerated dynamics~\eqref{eqn:sde-2} by
\begin{align}
  \hat{\vect b} = \alpha \vect b_s + \vect b_f,
\end{align}
and keeps the diffusion coefficient $\tilde\sigma$ unchanged.
Therefore, the difference between the observable calculated by the
accelerated dynamics and the original observable is given by 
\begin{align}\nonumber
  O(T) &= \tilde O(T/\alpha)\\\nonumber  
  &=\int \dd\vect y_0 {\rho}(\vect y_0, 0)
  \int_{\mc\{T/\alpha; \vect y_0, 0\}} 
  O[\vect y_t]\,\dd\tilde\pathmeas[\vect y_t]\\\nonumber
  &=
  \int \dd\vect y_0 {\rho}(\vect y_0, 0)
  \int_{\mc\{T/\alpha; \vect y_0, 0\}}
  O[\vect y_t] \frac{\dd\tilde\pathmeas[\vect y_t]}{\dd\hat\pathmeas[\vect y_t]}
  \dd\hat\pathmeas[\vect y_t]
\end{align}
Since we have $\tilde{\vect b} - \hat{\vect b} = (\alpha - 1)\vect b_f$,
by the Girsanov transformation, we have
\begin{align}\label{eqn:meas-reweight}
  \frac{\dd\tilde\pathmeas[\vect y_t]}{\dd\hat\pathmeas[\vect y_t]}
  = \exp
  \Big[
  \frac{\alpha-1}{\tilde{\sigma}}\int_0^{T/\alpha}\vect b_f(\vect y_s)\cdot\dd\vect w_s
  -
  \frac{(\alpha-1)^2}{2\tilde\sigma^2} \int_0^{T/\alpha}\vect b_f^2(\vect y_s)\dd t\,
  \Big]
\end{align}
We therefore have the error estimate
\begin{align}\label{eqn:o-error}
  O(T) = \hat O(T/\alpha)  +
  \int \dd\vect y_0 {\rho}(\vect y_0, 0)\cdot \me(\vect y_0, T/\alpha)
\end{align}
where
\begin{align}\label{eqn:error}
  \me(\vect y_0, t) = 
  \int_{\mc\{T/\alpha; \vect y_0, 0\}}
  O[\vect y_t]
  \cdot
  W[\vect y_t]\,\dd\hat\pathmeas
\end{align}
and $W[\vect y_s]$ is defined by
\begin{align}\label{eqn:error-exp}
  W[\vect y_t] = 
  \exp
  \Big[
  \frac{\alpha-1}{\tilde{\sigma}}\int_0^{T/\alpha}\vect b_f(\vect y_s)\cdot\dd\vect w_s
  -
  \frac{(\alpha-1)^2}{2\tilde\sigma^2} \int_0^{T/\alpha}\vect b_f^2(\vect y_s)\dd t\,
  \Big]
  -1
\end{align}
Now the problem is that when the accelerate
approximation~\eqref{eqn:o-error} has a good accuracy. One possibility
is that the measure reweighting~\eqref{eqn:meas-reweight} is small.
However, as a matter of fact, the reweighting is usually not small,
because the fast part of the force is usually large. The other possibility
is that the observable $O[\vect y]$ has vanished correlation with
$W[\vect y]$.
This is usually true if in the system the fast DOFs are decoupled with
the slow variables, and the observation is only on the space spanned
by the slow variables.

@@

The other fast variable is the repulsive part of the Lennard-Jones
interaction, which can be removed by the following way: we define a
smooth switching function $s(r)$ that has value 0 at repulsive part of
the LBJ potential, while has value of 1 at the attractive part. Then
the accelerated LJ potential (denoted by $\hat u(r)$) can be given by
\begin{align}
  \hat u(r) = \alpha s(r) u(r) + (1 - s(r)) u(r)
\end{align}
where $u(r)$ is the original LJ potential. It is clear that
$\hat u(0) = u(0)$, and $\hat u(\infty) = \alpha u(\infty)$.
\\

\recheck{ A possible idea is to estimate the error~\eqref{eqn:error}
  by the Girsanov transformation~\eqref{eqn:meas-reweight}. However, we
  suffer from the following difficulty: since the fast part of the
  force are generally very stiff, the magnitude of $\vect b_f$ is
  large comparing to the noise in the system,
  i.e.~$\tilde\sigma$. In our alanine dipeptide simulation,
  the typical magnitude of
  $\frac{\alpha-1}{\tilde{\sigma}}\int_0^{T/\alpha}\vect b_f(\vect
  y_s)\cdot\dd\vect w_s$ is of order $10^3$, while the magnitude of
  $\frac{(\alpha-1)^2}{2\tilde\sigma^2} \int_0^{T/\alpha}\vect
  b_f^2(\vect y_s)\dd t$ is of order $10^7$.  This means that for the
  trajectories we sampled, the reweighting~\eqref{eqn:meas-reweight}
  vanishes. There are also trajectories, along which the magnitude of
  the fast forces is very small (for example the bonds and angles stay
  around the equilibrium value), which means a significant contribution
  to the reweighting~\eqref{eqn:meas-reweight}, however  these
  trajectories are seldomly (actually not) sampled in real
  simulations.

  When we remove the fast forces from the system, actually the
  microscopic dynamics of the system changes a lot, that is why the
  direct application of Girsanov
  transformation~\eqref{eqn:meas-reweight} does not work.  However, we
  are only interested by the dynamics of the slow observables, which may NOT
  be perturbed a lot by removing the fast forces.  The question is how
  to derive an effective reweighting (or correction) for only the slow variables?

  If the fast varibles only contribute entropy to the slow degrees of
  freedom, and if during the nonequilibrium dynamics the fast varibles
  relaxes much faster than the slow variables (short-term entropy
  contribution is well defined), it may be possible to effectively
  add back the removed fast force (defined on fast DOFs)  by
  applying addtional free energy correction $\nabla_{\vect r}( \Delta A)$
  (on slow DOFs), 
  where $\Delta A$ denotes the free energy difference between
  the original system and the perturbed system (fast force removed).
  % then it may be possible to consider the Girsanov
  % transformation based on the free energy (on the slow variables)
  % difference:
%   \begin{align}\label{eqn:meas-reweight-da}
%     \frac{\dd\tilde\pathmeas[\vect y_t]}{\dd\hat\pathmeas[\vect y_t]}
%     = \exp
%     \Big[
%     \frac{\alpha-1}{\tilde{\sigma}}\int_0^{T/\alpha} \nabla( \Delta A)\cdot\dd\vect w_s
%     -
%     \frac{(\alpha-1)^2}{2\tilde\sigma^2} \int_0^{T/\alpha}\vert \nabla( \Delta A) \vert^2\dd t\,
%     \Big]
% \end{align}  
}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-01-check-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-01-check-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-01-check-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-01-check-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-01-check-5.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-5.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-5.eps}\\
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-2.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-3.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-4.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-5.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-3.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-4.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-5.eps}\\
  \caption{The integrated probability fluxes. $\alpha = 2,4,8$.}
  \label{fig:tmp3}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-01-check-1-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-01-check-1-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-01-check-1-3.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-1-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-1-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-02-1-3.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-04-1-3.eps}\\
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1-1.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1-2.eps}
  % \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-05-1-3.eps}\\
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1-1.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1-2.eps}
  \includegraphics[width=0.195\textwidth]{figs/fig-meta-iflux-08-1-3.eps}
  \caption{The integrated probability fluxes. $\alpha = 4,8$.}
  \label{fig:tmp3}
\end{figure}
